Introducing an Ecological Trait-data Standard

### 2017-11-14 manuscript/manuscript.Rmd
manuscript: minor note
fix URL and title

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane LeProvost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
output:
  word_document:
    fig_caption: yes
    fig_width: 6
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
bibliography: div05.bib
---

@@

Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13]. Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/).

@@

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/TraitDataStandard/.

@@

The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

@@

Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/TraitDataStandard/. 
Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. 
By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset. The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.

See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/TraitDataStandard/#terms-for-traitlists-a-trait-thesaurus].  We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

@@

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a set of terms (https://ecologicaltraitdata.github.io/TraitDataStandard/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.

@@

- the `occurrence` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age. This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/TraitDataStandard/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It aims at important information to track potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference).

@@

The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/fdschneider/traitdataform/issues).

>>>>

---
title: "Introducing an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane LeProvost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
output:
  word_document:
    fig_caption: yes
    fig_width: 6
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
bibliography: div05.bib
---

@@

Lists of trait definitions have been developed for plants, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html, @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait ontologies, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14], the Hymenoptera Anatomy Ontology [@yoder10, http://hymao.org/], or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html, @park13]. Hosting services for  biological ontologies and terminologies are the OBO foundry (http://obofoundry.org/) or ontobee (http://www.ontobee.org/). 
<!--General classes of traits are defined within the GeoBON Essential Biodiversity Variables [@pereira13].-->

@@

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/ETS/.

@@

The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v1.0/". In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

@@

Column names and, where possible, field entries should follow the defined vocabulary listed at https://ecologicaltraitdata.github.io/ETS/. 
Individual terms should be referenced via their global identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically (i.e. via the Application Programming Interface, API). Wherever our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

If no published trait definition is available that can be referenced, trait-datasets should be accompanied by a dataset-specific glossary of traits, or thesaurus. A thesaurus provides a "controlled vocabulary designed to clarify the definition and structuring of key terms and associated concepts in a specific discipline" [@laporte13; @garnier17]. 
Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. 
By providing a minimal vocabulary for this case (https://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus), we hope to facilitate the publication of trait thesauri developed for the own project context which always should be referenced in the core trait dataset. The thesaurus might accompany the core data in a Darwin Core Archive, or be published on any other stable webservice.

See our online reference for recommendations how to publish a minimal Thesaurus compliant with semantic web standards [http://ecologicaltraitdata.github.io/ETS/#terms-for-traitlists-a-trait-thesaurus].  We also provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github (http://ecologicaltraitdata.github.io/TraitDataList/).

@@

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a set of terms (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows to relate trait data to authorship and define a bibliographic reference and license model.

@@

- the `occurrence` extension (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence) contains vocabulary to describe information on the level of individual specimens, such as physiological specifications of the phenotype, such as sex, life stage or age. This also includes the date and geographical location, as well as the method of sampling and preservation, which provides an important resource to analyse trait variation due to differences in space and time.
- the `measurementOrFact` extension (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact) provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It aims at important information to track potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference).

@@

The package is under continuous open source development and invites participation in development, comments or bug reports via the Github Issue page (https://github.com/EcologicalTraitData/traitdataform/issues).
